---
title: "p8105_hw2_rx2236"
author: "Ruiqi Xue"
date: "2023-09-30"
output: github_document
---

```{r}
library(tidyverse)
library(readxl)
```


## Problem 1

Clean `pols` data

```{r}
month_df = 
  tibble(
    month_num = 1:12,
    month_abb = month.abb,
    month = month.name
  )

pols = 
  read_csv("./data/fivethirtyeight_datasets/pols-month.csv") |>
  separate(mon, into = c("year", "month_num", "day"), convert = TRUE) |>
  mutate(
    president = recode(prez_gop, "0" = "dem", "1" = "gop", "2" = "gop")) |>
  left_join(x = _, y = month_df) |> 
  select(year, month, everything(), -day, -starts_with("prez"))
```

Clean `snp` data

```{r}
snp = 
  read_csv("./data/fivethirtyeight_datasets/snp.csv") |>
  separate(date, into = c("month", "day", "year"), convert = TRUE) |>
  arrange(year, month) |>
  mutate(month = month.name[month]) |>
  mutate(year = ifelse(year >= 50, 1900 + year, 2000 + year)) |>
  select(year, month, close) 
```


Clean `unemployment` data

```{r}
unemployment = 
  read_csv("./data/fivethirtyeight_datasets/unemployment.csv") |>
  rename(year = Year) |>
  pivot_longer(
    Jan:Dec, 
    names_to = "month_abb",
    values_to = "unemployment"
  ) |> 
  left_join(x = _, y = month_df) |> 
  select(year, month, unemployment)
```


Merge three datasets

```{r}
data_538 = 
  left_join(pols, snp) |>
  left_join(x = _, y = unemployment)

str(data_538)
```

### Datasets explanation

The `pols` data has `r nrow(pols)` observations and `r ncol(pols)` variables and tells us about the party affiliation distribution (democrat or republican) for governors and senators for a given year from years `r range(pols$year)[1]` to `r range(pols$year)[2]`. It also tells us whether the sitting president was a democrat or republican. The `snp` data has `r nrow(snp)` observations and `r ncol(snp)` variables, ranging from years `r range(snp$year)[1]` to `r range(snp$year)[2]`. The `unemployment` data has `r nrow(unemployment)` observations and `r ncol(unemployment)` variables ranging from years `r range(unemployment$year)[1]` to `r range(unemployment$year)[2]`. In Januarys in or after 1975 in which a democrat was president, the **average unemployment rate was `r filter(data_538, month == "January", year >= 1975, president == "dem") |> pull(unemployment) |> mean() |> round(2)`**.  The average unemployment rate over the same time period in which a republican was president was `r filter(data_538, month == "January", year >= 1975, president == "gop") |> pull(unemployment) |> mean() |> round(2)`.



## Problem 2

Clean the Mr. Trash Wheel sheet

```{r clean the Mr. Trash Wheel sheet}
mr_trash_wheel = read_excel("./data/202207 Trash Wheel Collection Data.xlsx",
                            sheet = "Mr. Trash Wheel",
                            range = "A2:M549") |>
  rename_with(~ tolower(gsub(" ", "_", .x, fixed = TRUE))) |>
  rename(weight = "weight_(tons)", volume = "volume_(cubic_yards)") |>
  mutate(homes_powered = weight*500/30)
  

```


Clean the Professor Trash Wheel sheet

```{r clean the Professor Trash Wheel sheet}
professor_trash_wheel = read_excel("./data/202207 Trash Wheel Collection Data.xlsx",
                            sheet = "Professor Trash Wheel",
                            range = "A2:L96") |>
  rename_with(~ tolower(gsub(" ", "_", .x, fixed = TRUE))) |>
  rename(weight = "weight_(tons)", volume = "volume_(cubic_yards)") |>
  mutate(homes_powered = weight*500/30)
  

```


Clean the Gwynnda Trash Wheel sheet

```{r clean the Gwynnda Trash Wheel sheet}
gwynnda_trash_wheel = read_excel("./data/202207 Trash Wheel Collection Data.xlsx",
                            sheet = "Gwynnda Trash Wheel",
                            range = "A2:J108") |>
  rename_with(~ tolower(gsub(" ", "_", .x, fixed = TRUE))) |>
  rename(weight = "weight_(tons)", volume = "volume_(cubic_yards)") |>
  mutate(homes_powered = weight*500/30)


```





