---
title: "p8105_hw2_rx2236"
author: "Ruiqi Xue"
date: "2023-09-30"
output: github_document
---

```{r}
library(tidyverse)
library(readxl)
```


## Problem 1

Clean `pols` data

```{r}
month_df = 
  tibble(
    month_num = 1:12,
    month_abb = month.abb,
    month = month.name
  )

pols = 
  read_csv("./data/fivethirtyeight_datasets/pols-month.csv") |>
  separate(mon, into = c("year", "month_num", "day"), convert = TRUE) |>
  mutate(
    president = recode(prez_gop, "0" = "dem", "1" = "gop", "2" = "gop")) |>
  left_join(x = _, y = month_df) |> 
  select(year, month, everything(), -day, -starts_with("prez"))
```

Clean `snp` data

```{r}
snp = 
  read_csv("./data/fivethirtyeight_datasets/snp.csv") |>
  separate(date, into = c("month", "day", "year"), convert = TRUE) |>
  arrange(year, month) |>
  mutate(month = month.name[month]) |>
  mutate(year = ifelse(year >= 50, 1900 + year, 2000 + year)) |>
  select(year, month, close) 
```


Clean `unemployment` data

```{r}
unemployment = 
  read_csv("./data/fivethirtyeight_datasets/unemployment.csv") |>
  rename(year = Year) |>
  pivot_longer(
    Jan:Dec, 
    names_to = "month_abb",
    values_to = "unemployment"
  ) |> 
  left_join(x = _, y = month_df) |> 
  select(year, month, unemployment)
```


Merge three datasets

```{r}
data_538 = 
  left_join(pols, snp) |>
  left_join(x = _, y = unemployment)

str(data_538)
```

### Datasets explanation

The `pols` data has `r nrow(pols)` observations and `r ncol(pols)` variables and tells us about the party affiliation distribution (democrat or republican) for governors and senators for a given year from years `r range(pols$year)[1]` to `r range(pols$year)[2]`. It also tells us whether the sitting president was a democrat or republican. The `snp` data has `r nrow(snp)` observations and `r ncol(snp)` variables, ranging from years `r range(snp$year)[1]` to `r range(snp$year)[2]`. The `unemployment` data has `r nrow(unemployment)` observations and `r ncol(unemployment)` variables ranging from years `r range(unemployment$year)[1]` to `r range(unemployment$year)[2]`. In Januarys in or after 1975 in which a democrat was president, the **average unemployment rate was `r filter(data_538, month == "January", year >= 1975, president == "dem") |> pull(unemployment) |> mean() |> round(2)`**.  The average unemployment rate over the same time period in which a republican was president was `r filter(data_538, month == "January", year >= 1975, president == "gop") |> pull(unemployment) |> mean() |> round(2)`.



## Problem 2

Clean the Mr. Trash Wheel sheet

```{r clean the Mr. Trash Wheel sheet}
mr_trash_wheel = read_excel("./data/202309 Trash Wheel Collection Data.xlsx",
                            sheet = "Mr. Trash Wheel",
                            range = "A2:M586") |>
  rename_with(~ tolower(gsub(" ", "_", .x, fixed = TRUE))) |>
  rename(weight = "weight_(tons)", volume = "volume_(cubic_yards)") |>
  mutate(homes_powered = weight*500/30)
  

```


Clean the Professor Trash Wheel sheet

```{r clean the Professor Trash Wheel sheet}
professor_trash_wheel = read_excel("./data/202309 Trash Wheel Collection Data.xlsx",
                            sheet = "Professor Trash Wheel",
                            range = "A2:L108") |>
  rename_with(~ tolower(gsub(" ", "_", .x, fixed = TRUE))) |>
  rename(weight = "weight_(tons)", volume = "volume_(cubic_yards)") |>
  mutate(homes_powered = weight*500/30)
  

```


Clean the Gwynnda Trash Wheel sheet

```{r clean the Gwynnda Trash Wheel sheet}
gwynnda_trash_wheel = read_excel("./data/202309 Trash Wheel Collection Data.xlsx",
                            sheet = "Gwynnda Trash Wheel",
                            range = "A2:K157") |>
  rename_with(~ tolower(gsub(" ", "_", .x, fixed = TRUE))) |>
  rename(weight = "weight_(tons)", volume = "volume_(cubic_yards)") |>
  mutate(homes_powered = weight*500/30) |>
  mutate(month = str_to_title(month))

```


Add an additional variable to all datasets to keep track of which Trash Wheel is which 

```{r add a variable to all datasets}
mr_trash_wheel[, "trash_wheel"] = "Mr."
mr_trash_wheel = relocate(mr_trash_wheel, trash_wheel, .after = dumpster)

professor_trash_wheel[, "trash_wheel"] = "Professor"
professor_trash_wheel = relocate(professor_trash_wheel, trash_wheel, .after = dumpster)

gwynnda_trash_wheel[, "trash_wheel"] = "Gwynnda"
gwynnda_trash_wheel = relocate(gwynnda_trash_wheel, trash_wheel, .after = dumpster)

```


Combine these three datasets

```{r combine three datasets}
mr_trash_wheel = mutate(mr_trash_wheel, year = as.numeric(year))

combined_df = full_join(mr_trash_wheel, professor_trash_wheel) |>
  full_join(x = _, y = gwynnda_trash_wheel)

```

After combining the three datasets, the resulting dataset contains `r nrow(combined_df)` observations,
and `r ncol(combined_df)` variables. Key variables include `trash_wheel` which is the new-added variable indicating the name of Trash Wheel for each dumpster; `month`, `year` and `date` indicating the date of each dumpster; `weight` and `volume` indicating the weight and volume of each dumpster. There are several variables such as `plastic_bottles` indicating the types of trash and the corresponding amount of each dumpster. Finally, there is `homes_powered` indicating the approxiamate number of households the trash provides electricity for.



The total weight of trash collected by Professor Trash Wheel was `r sum(professor_trash_wheel$weight)`. 
And the total number of cigarette butts collected by Gwynnda in July of 2021 was `r sum(filter(gwynnda_trash_wheel, month == "July" & year == 2021)$cigarette_butts)`.



## Problem 3

Import and clean the baseline dataset

```{r clean the baseline dataset}
baseline_df = 
  read_csv("data/data_mci/MCI_baseline.csv",
           skip = 1) |>
  janitor::clean_names() |>
  rename("age_at_baseline" = "current_age") |>
  mutate(
    sex = case_match(
      sex,
      1 ~ "male",
      0 ~ "female")) |>
  mutate(
    apoe4 = case_match(
      apoe4,
      1 ~ "carrier",
      0 ~ "non-carrier"
    )
  )
```

There are `r ncol(baseline_df)` variables in this dataset. `id` identifies each participant, `education` indicates number of years of education, `apoe4` indicates the APOE4 status of the participants. `age_at_baseline` and `age_at_onset` indicate age relavant information accordingly. Up to this point we cleaned the dataset with proper variable names, and encoded the `sex` and `APOE4 carrier status` as non-numeric. From here, we see that `r nrow(baseline_df)` patients were recruited at the baseline.



```{r remove no MCI}
baseline_df  = 
  filter(baseline_df, age_at_baseline < age_at_onset | age_at_onset == ".")
  
```
Now we remove the participants who do not meet the stated inclusion criteria (i.e. no MCI at baseline).

As a result, we get the number of participants who developed MCI is `r nrow(baseline_df)`.

Calculated from the current dataframe, the average baseline age is `r mean(baseline_df$age_at_baseline)`. 


And the proportion of women in the study who are APOE4 carriers is `r nrow(filter(baseline_df, sex == "female" & apoe4 == "carrier")) / nrow(filter(baseline_df, sex == "female"))`.

Now import the amyloid dataset and clean the variable names.

```{r}
amyloid_df = 
  read_csv("data/data_mci/mci_amyloid.csv",
           skip = 1) |>
  janitor::clean_names()

```

At this point we have `r ncol(amyloid_df)` variables including a `study_id` and 5 time variables, and `r nrow(amyloid_df)` entries.


Now we pivot the dataframe.

```{r}
amyloid_df = amyloid_df |>
  pivot_longer(
    baseline:time_8,
    names_to = "time",
    values_to = "biomarker_value"
  ) 
  
```

After pivoting, the dataset has `r ncol(amyloid_df)` variables including `study_id`, `time` and `biomarker_value` which is the amyloid beta 42/40 ratio; and `r nrow(amyloid_df)` entries.


Now check whether some participants appear in only the baseline or amyloid datasets.

```{r}
only_in_baseline = subset(baseline_df, !id %in% amyloid_df$study_id)
print(only_in_baseline)

only_in_amyloid = subset(amyloid_df, !study_id %in% baseline_df$id)
print(only_in_amyloid)
```

So, there are `r nrow(only_in_baseline)` participants appear in only the baseline dataset, and `r n_distinct(only_in_amyloid$study_id)` participants appear in only the amyloid dataset. 


Now combine the two datasets so that only participants who appear in both datasets are retained using inner join.

```{r}
amyloid_df = rename(amyloid_df, id = study_id)
inner_combined = inner_join(baseline_df, amyloid_df)
```

We first to rename the `study_id` variable in the amyloid dataset into `id` as the same as in the baseline dataset so that the inner join can be performed with this common variable.

The combined dataset has `r nrow(distinct(inner_combined))`



